% Read STYLE.md before editing prose in this file.
% Follow the Content Creation Workflow in WORKFLOW.md §Content Creation Workflow.

\section{Methodology}
\label{sec:methodology}

\subsection{Data Collection}
\label{subsec:data-collection}

Acoustic recordings were obtained at three Atlantic sites: Site~A
(western Azores archipelago), Site~B (Bay of Biscay, northern Spain),
and Site~C (Cape Verde islands).
Field observations were conducted between June~2021 and September~2024,
yielding 42~confirmed departure events for which full acoustic records
were available.
A departure event was defined as a session in which a marked individual
or coherent group left the observation area and did not return within
the 72-hour monitoring window.

\begin{definition}[Departure Event]
	\label{def:departure-event}
	A \emph{departure event} is a continuous acoustic recording window of
	\(\pm 5\) minutes around the last observed surfacing of a focal
	individual or group, provided that individual or group did not re-enter
	the monitoring area within 72~hours of the recording.
\end{definition}

Recordings were made using calibrated hydrophone arrays deployed at
depths of 3--8~m, sampling at 96\kHz{} with 24-bit resolution.
All recordings were stored in lossless FLAC format and are archived
with the \citet{GuideResearch2025dataset} dataset.

\subsection{Annotation Scheme}
\label{subsec:annotation}

Each departure event was segmented into discrete acoustic units by two
independent annotators following the protocol in
\citet{GuideResearch2025dataset}.
Inter-annotator agreement, measured by Cohen's~$\kappa$, was $0.83$
across the full corpus---within the range typically considered
``substantial'' agreement~\citep{Dent2024farewell}.

Each unit was assigned one of five labels:
\texttt{whistle}, \texttt{burst-pulse}, \texttt{click-train},
\texttt{silence}, or \texttt{noise}.
The sequence of labels for a departure event constitutes a
\emph{vocalization trace}.

\subsection{Computational Analysis}
\label{subsec:analysis}

\paragraph{Hidden Markov Model}

We model each vocalization trace as a sequence emitted by a discrete-time
HMM with $N$ hidden states representing acoustic context:

\begin{equation}
	\label{eq:hmm-likelihood}
	P(\mathbf{o} \mid \lambda) =
	\sum_{\mathbf{q}} P(\mathbf{o} \mid \mathbf{q}, \lambda)\,
	P(\mathbf{q} \mid \lambda),
\end{equation}

where $\mathbf{o} = (o_1, \ldots, o_T)$ is the observed label sequence,
$\mathbf{q} = (q_1, \ldots, q_T)$ is the hidden state sequence, and
$\lambda = (A, B, \pi)$ denotes the transition matrix, emission matrix,
and initial distribution, respectively.

Model parameters were estimated using the Baum--Welch algorithm with
random restarts ($k=50$).
The number of hidden states $N$ was selected by Bayesian information
criterion over the range $N \in \{2, 3, \ldots, 10\}$; the
optimal value was $N = 4$.

\paragraph{Spectral Envelope Analysis}

For each annotated whistle unit, we extracted the spectral envelope using
linear predictive coding (LPC) with order $p = 16$.
The resulting envelope coefficients were projected onto a 2D space via
principal component analysis (PCA) for visualization.

\Cref{tab:vocalization-comparison} summarizes the acoustic features of
the three identified archetypes.

\input{tables/comparison_table}
